Unsupervised embeddings capture latent structure of scientific mobility

### 2020-02-27 paper/current/main.tex

% %Organization information
Affiliations mapped to one of 8,661 organizations, disambiguated following a mostly manual process originally designed for the Leiden Rankings of World Universities~\autocite{waltman2012leidenrankings}. 
Organizational records were associated with a full name, a type indicating the sector (e.g., University, Government, Industry), and an identifier for the country and city of the organization. 
Sixteen different sector types were included in the analysis, which we aggregated to a four high-level codes: \textit{University}, \textit{Hospital}, \textit{Government}, and \textit{Other}. 
Each record was also associated with a latitude and longitude, however for many organizations these were missing or incorrect. We manually updated the coordinates of 2,267 organizations by searching the institution name and city on Google Maps;
in cases where a precise location of the organization could not be identified, we used the coordinates returned when searching the name of the city.
The data was further enriched with country-level information, including continent, majority language spoken, and language family of the majority language (e.g., the language family of \textit{Spanish} is \textit{Italic}).  
State/province level  information were added using the reverse geocoding service LocationIQ using each organization's latitude and longitude as input. 
Regional census classifications were also added for states in the United States.
For each organization, we also calculated a size as the average number of unique authors (mobile and non-mobile) who published with that organization across each year of our dataset;
in the case that authors publish with multiple affiliations in a single year, they are counted towards each.

As a result of our disambiguation procedure, some affiliations are mapped to two organizations, one specific, and one more general.
For example, any author affiliated with ``Indiana University Bloomington'' will also be listed as being affiliated with the ``Indiana University System'', a more general designation for all public universities in Indiana. 
However, a more general organization may not always occur alongside the more specific one.
For example, a researcher affiliated with the smaller regional school ``Indiana University South Bend'' will be listed as affiliated with only the ``Indiana University System''. 
We identify all specific organizations that always co-occur along a more general one.
For every career trajectory that includes one of these specific organizations, we remove all occurrences of the more general organization;
trajectories containing only a general designation are not altered.

Scientific mobility is complex and can be defined in many possible ways~\autocite{robinson2019mobility}.
Here, we consider an author as mobile when at least two distinct organizational affiliations appear in their trajectory during the 2008-2019 time window. 
Under this definition, mobile authors constitute 3,007,192 or 23.2 percent of all authors, and 17,700,095 author-affiliation combinations. 
Mobile authors were associated with, on average, 2.5 distinct organizational affiliations. 
Rates of mobility tended to differ across country. 
For example, France, Qatar, the USA, Iraq, and Luxembourg had the most mobile authors (ref figure).
However, due to their size, the USA, accounted for nearly 40 percent of all mobile authors worldwide, with  10 countries accounting for 80 percent of all mobility (ref figure).
In most cases, countries with a high degree inter-organization mobility also had a high degree of international mobility;
however, the USA and France had a high degree of inter-organization mobility with the corresponding international mobility (ref figure).

@@

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling (SGNS) model for word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec} . 
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance. 
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory. 
Vectors are learned by training a neural network to optimize the prediction of so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory. 
The hidden layer of the trained neural network is used as the

%% Word2vec parameters and training procedure
The SGNS model has several tunable hyper-parameters, including the size of the learned vectors, the size of the token window  considered when training, the minimum frequency threshold for including an organization, and the number of iterations.
We iterate through several values of vector size and window size, repeating analysis for each combination. 
To mitigate the effect of less common organizations, only organizations appearing at least 50 times in mobility sentences are included in the embedding;
this yields 6,580, or 76 percent of all organizations.
We iterate the SGNS training for five epochs over all mobility sentences. 
All word embeddings are trained using the \textit{gensim} python implementation of the SGNS \textit{word2vec} model. 
We visualize embeddings by projecting into two dimensions using UMAP~\autocite{mcinnes2018umap}.

@@

Here, we apply SemAxis to U.S. universities by defining two axes, one of geography and one of academic prestige. 
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts. 
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Rankings or based on the mean normalized research impact sourced from the Leiden Rankings. 
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 universities top-ranked universities are selected and six of them are in the Northeastern United States, then the bottom six universities according to the same ranking will be included in the low-ranked subset. 
From the prestige axis we also derive what we call \textit{SemRank}, a university ranking based on the nearness of universities to the \textit{Elite} end of the axis. 
We compare the SemAxis ranking to the traditional university rankings using Spearman rank correlations.

@@

Results are reported for embeddings with a dimensionality of 300 and a window size of one, which were the parameters that produced  that best explained the flux between organizations (see Fig. SI~\ref{fig:supp:hyperparams}). 
We note that this performance and the outcomes of analysis was robust to hyperparemeters settings.

@@

Six country clusters were identified, reflecting a mixture of the regional, linguistic, historical clusters. 
From top-to-bottom, we identify a primarily East-Asian cluster (orange), a Scandinavian cluster (dark purple), a cluster of British Commonwealth countries (light purple), a central and eastern European cluster (sky blue), a diverse cluster of European and South American clusters (dark green), and finally another diverse cluster consisting of countries bordering the Mediterranean or the Black Sea.
When not defined by geography, clusters appear to be defined by linguistic and historical ties, as is the case for the commonwealth countries. 
By looking closely at the deprogram and meta-data, there is a tendency that countries are merged as a group if they share the language in the early stage of the deprogram and geographical aspect plays a more major role in the later stage of the dendrogram.

We investigate the extent to which geography and language dictate the formation of clusters using \textit{CluSim}~\autocite{gates2019element}.
\textit{CluSim} is a method that calculates the similarity between the cluster results and, for hierarchical clustering, can adjust the relative importance of membership at different levels of the hierarchy with scaling parameter $r$, acting likes ``zooming lens''.  
If $r$ is high, the similarity is based on the lower levels of the dendrogram, and when $r$ is low, similarity is based on higher levels dendrogram. 
In fig. \ref{fig:clustering}b, we report three cluster similarity curves (continent, language, language family) with varying scaling parameter $r$.
As expected, continent plays a major role at higher levels of clustering process (low $r$), language and language family explain the clustering process more at the lower levels (high $r$). 
This suggests that geography, then language, are important factors contributing to the structure of international scientific mobility.

%
% Figure - CLUSTERING
%
\begin{figure}[p!]
	\centering
	\label{fig:clustering}
	\includegraphics[width=\textwidth]{\figdir/Clustering/cluster_country.pdf}
	\caption{
		\textbf{Geography, then language, dictate international mobility.}
		\textbf{a.} Hierarchically clustered similarity matrix of country vectors, aggregated as the mean of all organization vectors within the country that has a more than 25 organizations, excluding the United States. 
		Color of matrix cells corresponds to similarity between country vectors.
		Color of country names corresponds to their cluster.
		Color of three cell rows separated from the matrix correspond to, from left to right, the continent of the country, the language family~\autocite{ethnologue}, and the individual language. 
		\textbf{b.}	\textit{CluSim} \autocite{gates2019element} results demonstrating the factors dictating hierarchical clustering, varying by scaling parameter, $r$. 
		Continent better explains the grouping of country vectors at higher levels of the clustering processes.
		Language family, and then majority spoken language, better explain the grouping of countries at more granular. 
	}
\end{figure}

>>>>

% %Organization information
Affiliations mapped to one of 8,661 organizations, disambiguated following a mostly manual process originally designed for the Leiden Rankings of World Universities~\autocite{waltman2012leidenrankings}.
Organizational records were associated with a full name, a type indicating the sector (e.g., University, Government, Industry), and an identifier for the country and city of the organization.
Sixteen different sector types were included in the analysis, which we aggregated to a four high-level codes: \textit{University}, \textit{Hospital}, \textit{Government}, and \textit{Other}.
Each record was also associated with a latitude and longitude, however, for many organizations, these were missing or incorrect. We manually updated the coordinates of 2,267 organizations by searching the institution name and city on Google Maps;
in cases where a precise location of the organization could not be identified, we used the coordinates returned when searching the name of the city.
The data was further enriched with country-level information, including continent, majority language spoken, and language family of the majority language (e.g., the language family of \textit{Spanish} is \textit{Italic}).
State/province-level information was added using the reverse geocoding service LocationIQ using each organization's latitude and longitude as input.
Regional census classifications were also added for states in the United States.
For each organization, we also calculated size as the average number of unique authors (mobile and non-mobile) who published with that organization across each year of our dataset;
in the case that authors publish with multiple affiliations in a single year, they are counted towards each.

As a result of our disambiguation procedure, some affiliations are mapped to two organizations, one specific, and one more general.
For example, any author affiliated with ``Indiana University Bloomington'' will also be listed as being affiliated with the ``Indiana University System'', a more general designation for all public universities in Indiana.
However, a more general organization may not always occur alongside the more specific one.
For example, a researcher affiliated with the smaller regional school ``Indiana University South Bend'' will be listed as affiliated with only the ``Indiana University System''.
We identify all specific organizations that always co-occur along with a more general one.
For every career trajectory that includes one of these specific organizations, we remove all occurrences of the more general organization;
trajectories containing only a general designation are not altered.

Scientific mobility is complex and can be defined in many possible ways~\autocite{robinson2019mobility}.
Here, we consider an author as mobile when at least two distinct organizational affiliations appear in their trajectory during the 2008-2019 time window.
Under this definition, mobile authors constitute 3,007,192 or 23.2 percent of all authors and 17,700,095 author-affiliation combinations.
Mobile authors were associated with, on average, 2.5 distinct organizational affiliations.
Rates of mobility tended to differ across the country.
For example, France, Qatar, the USA, Iraq, and Luxembourg had the most mobile authors (ref figure).
However, due to their size, the USA, accounted for nearly 40 percent of all mobile authors worldwide, with  10 countries accounting for 80 percent of all mobility (ref figure).
In most cases, countries with a high degree inter-organization mobility also had a high degree of international mobility;
however, the USA and France had a high degree of inter-organization mobility with the corresponding international mobility (ref figure).

@@

%% word2vec
These mobility sentences are used as input to the standard skip-gram negative sampling (SGNS) model for word embedding, commonly known as \textit{word2vec}~\autocite{mikolov2013word2vec} .
In natural-language processing, word embeddings are used to construct dense vector-space representations of words and phrases, in which distance relationships between words correspond to a notion of semantic distance.
By embedding co-affiliation trajectories, we aim to learn a dense vector for every organization where the distance between vectors relates to the tendency for two organizations to co-occur in a token window around the target term in the same career trajectory.
Vectors are learned by training a neural network to optimize the prediction of the so-called context term given a target, in our case using one organization to predict the others appearing in the same co-affiliation trajectory.
The hidden layer of the trained neural network is used as the

%% Word2vec parameters and training procedure
The SGNS model has several tunable hyper-parameters, including the size of the learned vectors, the size of the token window considered when training, the minimum frequency threshold for including an organization, and the number of iterations.
We iterate through several values of vector size and window size, repeating the analysis for each combination.
To mitigate the effect of less common organizations, only organizations appearing at least 50 times in mobility sentences are included in the embedding;
this yields 6,580, or 76 percent of all organizations.
We iterate the SGNS training for five epochs overall mobility sentences.
All word embeddings are trained using the \textit{gensim} python implementation of the SGNS \textit{word2vec} model.
We visualize embeddings by projecting into two dimensions using UMAP~\autocite{mcinnes2018umap}.

@@

Here, we apply SemAxis to U.S. universities by defining two axes, one of geography and one of academic prestige.
The poles of the geographic axis are defined as the mean vector of all vectors corresponding to organizations in California, and then the mean of all vectors of organizations in Massachusetts.
For the prestige axis, we define a subset of top-ranked universities according to either the Times World University Rankings or based on the mean normalized research impact sourced from the Leiden Rankings.
The other end of the prestige axis is the geographically-matched (according to census region) set of universities ranked at the bottom of these rankings.
For example, if 20 universities top-ranked universities are selected and six of them are in the Northeastern United States, then the bottom six universities according to the same ranking will be included in the low-ranked subset.
From the prestige axis, we also derive what we call \textit{SemRank}, a university ranking based on the nearness of universities to the \textit{Elite} end of the axis.
We compare the SemAxis ranking to the traditional university rankings using Spearman rank correlations.

@@

Results are reported for embeddings with a dimensionality of 300 and a window size of one, which was the parameters that produced that best explained the flux between organizations (see Fig. SI~\ref{fig:supp:hyperparams}).
We note that this performance and the outcomes of the analysis were robust to hyperparameters settings.

@@

Six country clusters were identified, reflecting a mixture of the regional, linguistic, historical clusters.
From top-to-bottom, we identify a primarily East-Asian cluster (orange), a Scandinavian cluster (dark purple), a cluster of British Commonwealth countries (light purple), a central and eastern European cluster (sky blue), a diverse cluster of European and South American clusters (dark green), and finally another diverse cluster consisting of countries bordering the Mediterranean or the Black Sea.
When not defined by geography, clusters appear to be defined by linguistic and historical ties, as is the case for the commonwealth countries.
By looking closely at the dendrogram and meta-data, there is a tendency that countries are merged as a group if they share the language in the early stage of the deprogram and geographical aspect plays a more major role in the later stage of the dendrogram.

We investigate the extent to which geography and language dictate the formation of clusters using \textit{CluSim}~\autocite{gates2019element}.
\textit{CluSim} is a method that calculates the similarity between the cluster results and, for hierarchical clustering, can adjust the relative importance of membership at different levels of the hierarchy with scaling parameter $r$, acting likes ``zooming lens''.
If $r$ is high, the similarity is based on the lower levels of the dendrogram, and when $r$ is low, the similarity is based on higher levels of the dendrogram.
In fig. \ref{fig:clustering}a, we report three cluster similarity curves (continent, language, language family) with varying scaling parameter $r$.
As expected, continent plays a major role at higher levels of clustering process (low $r$), language and language family explain the clustering process more at the lower levels (high $r$).
This suggests that geography, then language, are important factors contributing to the structure of international scientific mobility.

%
% Figure - CLUSTERING
%
\begin{figure}[p!]
    \centering
    \label{fig:clustering}
    \includegraphics[width=\textwidth]{\figdir/Clustering/cluster_country.pdf}
    \caption{
        \textbf{Geography, then language, dictate international mobility.}
        \textbf{a.} Hierarchically clustered similarity matrix of country vectors, aggregated as the mean of all organization vectors within the country that has more than 25 organizations, excluding the United States.
        The color of matrix cells corresponds to the similarity between country vectors.
        Color of country names corresponds to their cluster.
        Color of three cell rows separated from the matrix correspond to, from left to right, the continent of the country, the language family~\autocite{ethnologue}, and the individual language.
        \textbf{b.}    \textit{CluSim} \autocite{gates2019element} results demonstrating the factors dictating hierarchical clustering, varying by scaling parameter, $r$.
        Continent better explains the grouping of country vectors at higher levels of the clustering processes.
        Language family, and then majority spoken language, better explain the grouping of countries at more granular.
    }
\end{figure}
